Northern Pikeminnow Predation on Juvenile Chinook Salmon in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Populations

### 2021-12-30 analysis/paper/paper.Rmd
added last references

We performed gastric lavage on nearly all Northern Pikeminnow collected during this study. In addition to juvenile Chinook Salmon, species included Redside Shiner, Largescale Sucker *Catostomus macrocheilus*, sculpin *Cottus* spp., and Mountain Whitefish *Prosopium williamsoni*; although decomposition from digestion rendered many fish contents unidentifiable. While this approach provided useful information on prey species, we have concerns about using angling and gastric lavage to quantify diet composition in Northern Pikeminnow. First, previous studies have indicated that diets of angled fish differ from those collected by other methods (e.g., electrofishing, [@Hodgson1998]), and further, angled fish are much more likely to have empty stomachs [@Jurajda2016]. Second, unlike other predators such as Smallmouth Bass and Walleye that are commonly sampled using gastric lavage [@Kamler2001], Northern Pikeminnow are cyprinids and therefore lack a true stomach. Previous research indicates that performing gastric lavage on cyprinid species can be ineffective at quantifying diet composition [@Hartleb1995], or alternately, requires a modification in methodology to flush food items out the vent of the fish [@Wasowicz1994]. Our intent, rather, was to simply document predation on juvenile Chinook Salmon in Deadwater Slough, and given the potential ineffectiveness of gastric lavage using angling and for Northern Pikeminnow, we consider our diet composition estimates to be conservative estimates i.e., we believe that a larger proportion of the Northern Pikeminnow diet is fish and juvenile Chinook Salmon than our results suggest. 
<!--
Although we only identified fish or fish parts within `r round(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"] * 100, 1)`% of Northern Pikeminnow examined and that fish remnants were only `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of the overall stomach contents, we believe those numbers to be misleading...
-->

@@

Our approach assumed that all available prey consumed by Northern Pikeminnow have energy densities equal to juvenile Chinook Salmon. Other salmonid including juvenile steelhead and Sockeye Salmon are likely also available prey in Deadwater Slough, especially during the spring migration. Additionally, hatchery smolt releases (Chinook Salmon, steelhead, and Sockeye Salmon) are also prevalent in the mainstem Salmon River during the spring migration. Other studies have used a similar approach where the energy densities of different salmonids were generalized using an estimate from a single species [@Peterson1999].<!-- The following needs to be updated after we finalize some results --> However, our study is focused on potential impacts to natural-origin DSR and NRR Chinook Salmon, in which case, let's consider a thought exercise where DSR and NRR Chinook Salmon represent only 10% and 25% of total fish prey consumed by Northern Pikemnnow during these periods. Under these scenarios, the average Northern Pikeminnow would consume 8.95 g (range 4.52 – 10.05 g) if 10% of prey were natural-origin Chinook Salmon and 22.38 g (range 11.31 – 25.12) if 25%.

>>>>

We performed gastric lavage on nearly all Northern Pikeminnow collected during this study. In addition to juvenile Chinook Salmon, species included Redside Shiner, Largescale Sucker *Catostomus macrocheilus*, sculpin *Cottus* spp., and Mountain Whitefish *Prosopium williamsoni*; although decomposition from digestion rendered many fish contents unidentifiable. While this approach provided useful information on prey species, we have concerns about using angling and gastric lavage to quantify diet composition in Northern Pikeminnow. First, previous studies have indicated that diets of angled fish differ from those collected by other methods (e.g., electrofishing, [@Hodgson1988]), and further, angled fish are more likely to have empty stomachs [@Jurajda2016], perhaps because they are hungry and actively searching for food. Second, unlike other predators such as Smallmouth Bass and Walleye that are commonly sampled using gastric lavage [@Kamler2001], Northern Pikeminnow are cyprinids and therefore lack a true stomach. Previous research indicates that performing gastric lavage on cyprinid species can be ineffective at quantifying diet composition [@Hartleb1995], or alternately, requires a modification in methodology to flush food items out the vent of the fish [@Wasowicz1994]. Our intent, rather, was to simply document predation on juvenile Chinook Salmon in Deadwater Slough, and given the potential ineffectiveness of gastric lavage using angling and for Northern Pikeminnow, we consider our diet composition estimates to be conservative estimates i.e., we believe that a larger proportion of the Northern Pikeminnow diet is fish and juvenile Chinook Salmon than our results suggest. 
<!--
Although we only identified fish or fish parts within `r round(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"] * 100, 1)`% of Northern Pikeminnow examined and that fish remnants were only `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of the overall stomach contents, we believe those numbers to be misleading...
-->

@@

Our approach assumed that all available prey consumed by Northern Pikeminnow have energy densities equal to juvenile Chinook Salmon. Other salmonid including juvenile steelhead and Sockeye Salmon are likely also available prey in Deadwater Slough, especially during the spring migration. Additionally, hatchery smolt releases (Chinook Salmon, steelhead, and Sockeye Salmon) are also prevalent in the mainstem Salmon River during the spring migration. Other studies have used a similar approach where the energy densities of different salmonids were generalized using an estimate from a single species [@Petersen1999].<!-- The following needs to be updated after we finalize some results --> However, our study is focused on potential impacts to natural-origin DSR and NRR Chinook Salmon, in which case, let's consider a thought exercise where DSR and NRR Chinook Salmon represent only 10% and 25% of total fish prey consumed by Northern Pikemnnow during these periods. Under these scenarios, the average Northern Pikeminnow would consume 8.95 g (range 4.52 – 10.05 g) if 10% of prey were natural-origin Chinook Salmon and 22.38 g (range 11.31 – 25.12) if 25%.
